\replaced[id=siyu]{Separating random and deterministic sources of computational noises in explore-exploit decisions}{The nature of decision noise in random exploration}

### 2022-12-19 manuscript/2022 draft - Plos Comp Bio/manuscript_DeterministicRandomNoise.tex
draft to bob

\newcommand{\specialcell}[2][c]{%
	\begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}
\setlength{\textheight}{9.3in}
\setlength{\textwidth}{7in}
\setlength{\footskip}{0.5in}
\setlength{\topmargin}{-0.5in}
\setlength{\headheight}{0.2in}
\setlength{\headsep}{0in}
\setlength{\parindent}{1pc}
\setlength{\oddsidemargin}{-0.25in}
\setlength{\evensidemargin}{-0.25in}
\renewcommand{\baselinestretch}{1.5}

%\usepackage[final]{changes}
\usepackage{changes}
\definechangesauthor[name={Siyu Wang}, color=red]{siyu}

\title{The nature of decision noise in random exploration}

\author[1,\textcurrency]{Siyu Wang}
\author[1,2,3]{Robert C. Wilson}

\affil[1]{Department of Psychology, University of Arizona, Tucson AZ, USA}
\affil[2]{Neuroscience and Physiological Sciences Graduate Interdisciplinary Program, University of
	Arizona, Tucson AZ, USA}
\affil[3]{Cognitive Science Program, University of Arizona, Tucson AZ, USA}
\affil[ \textcurrency]{\added[id=siyu]{Current Address: Laboratory of Neuropsychology, National Institute of Mental Health, National Institutes of Health, Bethesda MD, USA}}

@@

	\newpage
	\begin{abstract}

Human decision making is inherently variable. While this variability is often seen as a sign of suboptimality behavior, recent work suggests that variability can actually be adaptive. An example arises when we must choose between exploring unknown options or exploiting options we know well. A little randomness in these `explore-exploit' decisions is remarkably effective as it encourages us to explore options we might otherwise ignore. Recent work suggests that people may actually use such `random exploration' in practice, increasing their behavioral variability when it is more valuable to explore. 
%Other work, however, suggests that variability in human behavior is not always random, and is instead driven by deterministic factors (such as repeating past actions) 
A key question is whether the variability in random exploration is actually random. That is, is random exploration driven by stochastic processes in the brain or by some unobserved deterministic process that we have failed to account for when measuring behavioral variability? By designing an explore-exploit task in which, unbeknownst to them, participants are presented with the exact same choice twice, we provide a partial answer to this question. In particular, we find evidence that at least part \deleted[id=siyu]{27$\%$} of the variability in `random' exploration can be accounted for by deterministic processing of the stimulus. This still leaves open the possibility that much of random exploration is truly `random,' but narrows the window of opportunity for stochastic processes to explain this behavior.
%suggests that at least some `random' exploration is not random, but leaves open the possibility that the 
%However, it is unclear whether this variability is truly random
%From a modeling perspective, behavioral variability is essentially the variance that can not be explained by a model and is modeled as the level of decision noise. However, what we have called "decision noise" in previous researches could actually just be missing deterministic components from the model, it is difficult to tell whether decision noise truly arises from a stochastic process. Here we show that, while both random and deterministic noise drive variability in behavior, the noise driving random exploration is predominantly random. This suggests that random exploration depends on adaptive noise processes in the brain which are subject to cognitive control.

@@

	%BOB SAYS: IN GENERAL, NEED TO GO THROUGH THE PAPER AND MAKE SURE THAT `RANDOM' NOISE IS REPLACES WITH `NOT A DETERMINISTIC FUNCTION OF THE STIMULUS' OR SOMETHING LIKE THAT

	%(BOB SAYS: TAKE OR LEAVE THIS CHANGE TO DATE, BUT SETTING UP AS A DATE ALLOWS YOU TO TALK ABOUT AVOID YOUR EX LATER ON, WHICH MAY BE A BETTER EXAMPLE OF A DETERMINISTIC PROCESS!).
	Imagine trying to decide where to go to dinner on a date, you can go to your favorite restaurant, the one you both really enjoy and always go to, or you can try a new restaurant that you know nothing about. Such decisions, in which we must choose between a well-known `exploit' option and a lesser known `explore' option, are known as explore-exploit decisions.  From a theoretical perspective, making optimal explore-exploit choices, i.e. choices that maximize long-term reward, is computationally intractable in most cases \citep{eegittins74, eeBasu18}. In part because of this difficulty, there is considerable interest in how humans and animals solve the explore-exploit dilemma in practice \citep{, eeauer02,eegittins79,eekrebs78,eethompson33, eewatkins89, eebridle90,eemeyer95, eebanks97,eefrank09, eesteyvers09, eelee11, eepl12,eezhang13,eedaw06, eepl11, wilson2014}.

	One particularly effective strategy for solving the explore-exploit dilemma is choice randomization \citep{eethompson33, eewatkins89, eebridle90}. In this strategy, the decision process between exploration and exploitation is corrupted by `decision noise,' meaning that high value `exploit' options are not always chosen and exploratory choices are sometimes made by chance. In theory, such `random exploration,' is surprisingly effective and, if implemented correctly, can come close to optimal performance \citep{eethompson33, eebridle90, eeAgrawal11, eeChapelle11}.

	It has recently been shown that humans appear to use random exploration and can increase such decision noise when it is more beneficial to explore \citep{Gershman2018, wilson2014}. In one of these tasks, known as the Horizon Task, the key manipulation is the horizon condition, i.e. the number of decisions remaining for the participant to make. Increasing the horizon makes exploration more valuable as there is more time to use the information gained by exploration to maximize future rewards. For example, if you are leaving town tomorrow (short horizon), you will probably exploit the restaurant you know and love, but if you are in town for a while (long horizon), you would be more likely to explore the new restaurant. Using such a horizon manipulation we found that people's behavior is more variable in long horizons than short horizons, suggesting that they use adaptive decision noise to solve the explore-exploit dilemma \citep{wilson2014}.

	One limitation of this previous work, is that it is difficult to tell whether what is measured as decision noise is truly random. That is, whether behavioral variability is due to intrinsic stochastic processes in the brain or whether it is due to deterministic processes that we simply have not observed.  For example, in the restaurant example, my usual preference for one restaurant or another may be overruled if I see an ex romantic partner going into one of them. Avoiding an ex is a deterministic process, but if we fail to take ex's presence into account as scientists modeling the decision, then over a series of such decisions where the ex is present or not, we would mistakenly attribute the ensuing `variability' in choice to randomness.

	%This difficulty arises in how noise is defined in these previous tasks. For example, in the Horizon Task, variability is defined as the inverse temperature parameter of a softmax choice function.

	%In particular, in most cases noise is defined by fitting a softmax decision rule to behavior.

	%because noise in previous work is defined as the component of the decision process that cannot be accounted for by a model. 
	%, what we have called 'noise' in previous research could actually just be some missing deterministic components from the model. Decision noise as defined in previous researches are more or less a quantification of what's not predictable by the model. 
	%For example, in the restaurant case above, if you happen to spot an old friend walking in to one of the restaurants, you may be more likely to follow them in. Such a process would be deterministic, in that it is driven by an observable stimulus, but unless a model of your behavior incorporated the presence of your friend,

@@

	In this paper, we investigate the extent to which the apparent randomness in random exploration can be explained by such deterministic processing of the stimulus. In particular, we modify the Horizon Task to have people face the exact same explore-exploit choice twice. If the decision is a purely deterministic function of the stimulus, then people's choices should be identical both times. That is, their choices should be consistent, since the stimulus is the same both times. Conversely, the more their decision is driven by other processes, including both stochastic and unobserved deterministic processes, the less consistent their behavior should be. By analyzing behavior on this task in both a model-free and model-based manner, we show that at least some of the `randomness' in random exploration must come from  deterministic processing of the stimulus.  This does not prove that random exploration is entirely deterministic or stochastic, but provides a lower bound on how much deterministic processes contribute to random exploration. This in turn sheds light on how random exploration may be implemented by amplifying the effects of task-irrelevant stimuli on choice.

@@

	80 participants (ages 18-25, 37 male, 43 female) from the University of Arizona undergraduate subject pool participated in the experiment. 14 were excluded on the basis of performance, using the same exclusion criterion as in \citep{wilson2014}. In this exclusion criteria, we measured the accuracy of each participant's choices by calculating the percentage of times that a participant chose the bandit with the higher underlying mean payouts in the last choice of a long horizon game, intuitively people should figure out which bandit has a higher mean payout by the last trial and should have an accuracy measure significantly above 50\%, specifically, we computed the likelihood that the measured accuracy can be achieved driven by making a completely random choice between the two options and excluded participants with a likelihood smaller than 99.999\%, in other words, participants who didn't show an accuracy significant above chance with $p < 0.001$ were excluded in the analysis. This left 66 for the main analysis. Note that including the 14 badly performing subjects did not change the main results (Supplementary Figures 1 - 3)% MAKE AND INCLUDE REPLICAS OF ALL FIGS BUT WITH ALL SUBJECTS

		\begin{figure}[hp]
		\begin{center}
			\includegraphics[width=\textwidth]{figures/taskfiga.pdf}
			\caption{ 
			Schematic of the experiment. (A) Dynamics of an example horizon 6 game.  Here the first four trials are forced trials in which participants are instructed which option to play.  After the forced trials, participants are free to choose between the two options for the remainder of the game.  (B) Different possible states of the game after the first free choice over the course of the experiment. On average, participants play more than 150 such games, with varying horizon (1 vs 6), uncertainty condition ([1 3] vs [2 2]) and observed rewards.  In addition, all games are repeated (as Game 18 and 100 are here) such that participants will be faced with the exact same pattern of forced trials and exact same outcomes from those forced trials twice within each experiment.  These repeated games allow us to compute the relative contribution of deterministic and random noise by analyzing the extent to which choices are {\em consistent} across the repeated games.}
			\label{fig:taskfig}
		\end{center}
	\end{figure}

	The first four trials of each game were forced-choice trials, in which only one of the options was available for the participant to choose. We used these forced-choice trials to manipulate the relative ambiguity of the two options, by providing the participant with different amounts of information about each bandit before their first free choice. The four forced-choice trials set up two uncertainty conditions: unequal uncertainty(or [1 3]) in which one option was forced to be played once and the other three times, and equal uncertainty(or [2 2]) in which each option was forced to be played twice. After the forced-choice trials, participants made either 1 or 6 free choices (two horizon conditions),  Figure \ref{fig:taskfig}.

	\subsection*{Model-based analysis}
	We modeled behavior on the first free choice of the Horizon Task using a version of the logistic choice model in \citep{wilson2014} that was modified to differentiate between components of the noise that are deterministically driven by the stimulus (`deterministic noise') and components of the noise that are not deterministically driven by the stimulus (`random noise').  Because the stimuli are identical in the repeated games, by definition, deterministic noise remains the same in repeated games, whereas random noise can change.

	To model participants' choices on this first free-choice trial, we assume that they make decisions by computing the difference in value $\Delta Q$ between the right and left options, choosing right when $\Delta Q > 0$ and left otherwise.  Specifically, we write
	\begin{equation}
	\Delta Q= \Delta R+A \Delta I+b+n_{det}+n_{ran}
	\end{equation}
	where, the experimentally controlled variables are $\Delta R=R_{right}-R_{left}$, the difference between the mean of the rewards shown on the forced trials, and $\Delta I$, the difference of information available for playing the two options on the first free-choice trial. For simplicity, and because information is manipulated categorically in the Horizon Task, we define $\Delta I$ to be +1, -1 or 0, +1 if one reward is drawn from the right option and three are drawn from the left in the [1 3] condition, -1 if one from the left and three from the right, and in [2 2] condition, $\Delta I$ is 0.

	The other variables are: the spatial bias, $b$, which determines the extent to which participants prefer the option on the right; the information bonus $A$, which controls the level of directed exploration; $n_{det}$ and $n_{ran}$ are deterministic noise and random noise respectively. $n_{det}$ denotes the deterministic noise, which is identical on the repeat versions of each game; and $n_{ran}$ denotes random noise, which is uncorrelated between repeat plays and changes every game.

	\subsubsection*{Parameter recovery\label{ch:appendix:bayesrecovery}}

	To be sure that our fit parameter values were meaningful, we tested the ability of our model fitting procedure to recovery parameters from simulated data.  In particular, we simulated choices with the fitted parameters from the Hierarchical Bayesian analysis, and the re-fit the simulated choices to see whether we can recover the parameters.

	Results of this parameter recovery procedure are shown in Figure \ref{fig:pararecover1}. Parameter recovery is good (in terms of correlation between the true vs fitted parameters) for all parameters apart from the bias in short horizon condition, which is likely due to this parameter being so close to zero (Figure \ref{fig:pararecover1} Panel C). The recovery for the noise parameters, $\sigma_{det}$ and $\sigma_{ran}$, is slightly better for horizon 1 than horizon 6. This is because it requires more trials to recover larger values of $\sigma_{det}$ and $\sigma_{ran}$, so with the same number of choices it is harder to recover overall larger noise variances in horizon 6. In addition we see better recovery for random noise than deterministic noise. This is likely because we effectively have half as many trials for deterministic noise. In particular, while we generate two samples of random noise for each repeated game pair, we only generate one sample of deterministic noise, which by definition is the same in both of the repeated games.

	\begin{figure}[hp]
		\begin{center}
			\includegraphics[width=1\textwidth]{figures/precover.png}
			\caption{Parameter recovery over the subject-level means of information bonus, $A$, spatial bias, $b$, random noise variance, $\sigma_{ran}$, and deterministic noise variance, $\sigma_{det}$, for horizon 1 (left column) and horizon 6 (right column) games. 
			}
			\label{fig:pararecover1}
		\end{center}
	\end{figure}

	\begin{figure}[hp]
		\begin{center}
			\includegraphics[width=1\textwidth]{figures/RDBayes_parameterrecovery_grid_pureRanDet.jpg}
			\caption{Parameter recovery over the posterior of random noise variance, $\sigma_{ran}$, and deterministic noise variance, $\sigma_{det}$, for purely random noise (left column) and purely deterministic noise (right column) games. 
			}
			\label{fig:pararecover2}
		\end{center}
	\end{figure}

	Finally, despite the strong correlation between simulated and fit values of deterministic noise, we observed a systematic underestimation of $\sigma_{det}$. This is further illustrated in Supplementary Figure S5. In the simulation with 0 random noise and fully deterministic noise, our model perfectly recovered both random and deterministic noise, however in the simulation with full random noise and 0 deterministic noise, our model recovered only XXX \% of deterministic noise, and the remaining shows up as random noise instead (Figure \ref{fig:pararecover2}). Thus, our model provides a lower bound on deterministic noise and an upper bound on random noise.

	%Overall, we are able to recover both deterministic and random noises using our model to a satisfactory extent.

@@

	\subsection*{The Repeated-Games Horizon Task}
	We used a modified version of the `Horizon Task' \citep{wilson2014} to show the influence of stimulus-driven `deterministic noise' vs non-stimulus driven `random noise' on people's decisions (Figure \ref{fig:taskfig}). In this task, participants make repeated choices between two slot machines, or `one-armed bandits,' that pay out probabilistic rewards. Because they are initially unsure as to the mean payoff of each bandit, this task requires that participants carefully balance exploration of the lesser known bandit with exploitation of the better known bandit to maximize their overall rewards.

@@

	Relative to the original Horizon Task, the key modification here is to give people `repeated games,' in which they see exact same set of example plays twice in two separate games (separated by several minutes in time so as to avoid detection). By repeating the instructed plays for each game twice, we can set up a situation where (unbeknownst to the participants) they are faced with the exact same explore-exploit choice, with the exact same stimuli twice. Thus, if their behavior is a deterministic function of the stimuli, they will behave identically on the two games, that is their behavior on the two versions of each game will be consistent. Conversely, if their behavior is not driven by a deterministic function of the stimulus, then their choices on the repeated games should be inconsistent.

	On average participants played 67.31 repeated games each allowing us to quantify the extent to which their behavior was a deterministic function of the stimulus or not.

@@

	Before discussing the results for repeated games, we first confirm that the basic behavior in this task is consistent with our previously reported results \citep{wilson2014}. As in our previous work, we find evidence for two types of exploration in the Horizon Task.  Random exploration, which is the main focus of this paper, where exploration is driven by noise, and directed exploration, where exploration is driven by information.

	Random exploration is quantified in a model-free way as the probability of choosing the low mean option, $p(\mbox{low mean})$ in the equal, or [2 2], condition. This value increases with horizon, consistent with the idea that behavior is more random in horizon 6 (t(65) = 6.60, p $<$ 0.001 for [1 3], t(65) = 8.14, p $<$ 0.001 for [2 2]).  Directed exploration, is measured as the probability of choosing the more informative option $p(\mbox{high info}$ in the unequal, or [1 3], condition. Again this measure increases with horizon, showing that people are more information seeking in horizon 6 (t(65) = 7.00, p $<$ 0.001).

@@

	Next we asked whether participants' choices were consistent or inconsistent in the two repetitions of each game.  The idea behind this measure is that purely deterministic noise should lead to consistent choices as the deterministic stimulus is identical both times. Conversely, if choice is not a deterministic function of the stimulus, participants' choices should be independent, and hence more inconsistent across the repetitions of the game.

	To quantify choice inconsistency we computed the frequency with which participants made different responses for pairs of repeated games (Figure \ref{fig:mf2}). Using this measure we found that participants made inconsistent choices in both the unequal ([1 3]) and equal ([2 2]) information conditions, suggesting that not all of the noise was stimulus driven (t-test vs zero revealed that inconsistency was greater than zero for all horizon and uncertainty conditions.  For [1 3] condition, t(65) = 12.92, p $<$ 0.001 for horizon 1, t(65) = 16.76, p $<$ 0.001 for horizon 6; For [2 2] condition, t(65) = 9.67, p $<$ 0.001 for horizon 1, t(65) = 17.74, p $<$ 0.001 for horizon 6). In addition, we found that choice inconsistency was higher in horizon 6 than in horizon 1 for both [1 3] and [2 2] condition (F(1, 196) = 61.19, p $<$ 0.001), suggesting that at least some of the horizon dependent noise is not a deterministic function of the stimulus.

	\begin{figure}[h]
		\begin{center}
			\includegraphics[width=\textwidth]{figures/RanDetNoise_pinconsistent.jpg}
			\caption{Model-free analysis suggests that both deterministic and random noise contribute to the choice variability in random exploration. For both the [1 3] (A) and [2 2] (B) condition, people show greater choice inconsistency in horizon 6 than horizon 1. However, the extent to which their choices are inconsistent lies between what is predicted by purely deterministic and random noise, suggesting that both noise sources influence the decision.}
			\label{fig:mf2}
		\end{center}
	\end{figure}

>>>>

\newcommand{\specialcell}[2][c]{%
	\begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}
\setlength{\textheight}{9.3in}
\setlength{\textwidth}{7in}
\setlength{\footskip}{0.5in}
\setlength{\topmargin}{-0.5in}
\setlength{\headheight}{0.2in}
\setlength{\headsep}{0in}
\setlength{\parindent}{1pc}
\setlength{\oddsidemargin}{-0.25in}
\setlength{\evensidemargin}{-0.25in}
\renewcommand{\baselinestretch}{1.5}
\usepackage{float}

% control the fontsize
\PassOptionsToPackage{textsize=scriptsize}{todonotes}
%\usepackage[final]{changes}
\usepackage{changes}
\definechangesauthor[name={Siyu Wang}, color=red]{siyu}

\usepackage{xpatch}

%\usepackage{unravel}
%\unravelsetup{max-action=1000, max-input=1000, max-output=1000}
\providecommand\unravel[1]{#1}

% auto insert "comment"
\makeatletter
\xpatchcmd\replaced
{\setkeys{Changes@replaced}{#1}}
{\setkeys{Changes@replaced}{comment={(was): \protect\truncateto{10}{#3}}, #1}}
{}{}
\xpatchcmd\deleted
{\setkeys{Changes@deleted}{#1}}
{\setkeys{Changes@deleted}{comment={(deleted): \protect\truncateto{10}{#2}}, #1}}
{}{}

% fix \listofchanges
\xpatchcmd\ChangesListline
{0px}
{0pt}
{}{}

\newcounter{truncate}
\newcounter{truncate@max}

% truncate #2 to leave at most first #1 words
\def\truncateto#1#2{%
	\setcounter{truncate}{0}%
	\setcounter{truncate@max}{#1}%
	\truncate@loop{#2}%
}

\def\truncate@loop#1{%
	\ifnum\c@truncate=\c@truncate@max
	\def\next{[...]}% tail
	\else
	\in@{ }{#1}%
	\ifin@
	\stepcounter{truncate}%
	\truncate@split#1\@nil
	\else
	\def\next{#1}%
	\fi
	\fi
	\next
}

\def\truncate@split#1 #2\@nil{%
	\def\next{#1 \truncate@loop{#2}}%
}
\makeatother

\makeatletter
\setdeletedmarkup{\@gobble{#1}}
\makeatother

\title{\replaced[id=siyu]{Separating random and deterministic sources of computational noises in explore-exploit decisions}{The nature of decision noise in random exploration}}
\author[1,\textcurrency]{Siyu Wang}
\author[1,2,3]{Robert C. Wilson}

\affil[1]{Department of Psychology, University of Arizona, Tucson AZ, USA}
\affil[2]{Neuroscience and Physiological Sciences Graduate Interdisciplinary Program, University of
	Arizona, Tucson AZ, USA}
\affil[3]{Cognitive Science Program, University of Arizona, Tucson AZ, USA}
\affil[ \textcurrency]{Current Address: Laboratory of Neuropsychology, National Institute of Mental Health, National Institutes of Health, Bethesda MD, USA}

@@

	\newpage
	\begin{abstract}
Human decision making is inherently variable. While this variability is often seen as a sign of suboptimality behavior, recent work suggests that variability can actually be adaptive. An example arises when we must choose between exploring unknown options or exploiting options we know well. A little randomness in these `explore-exploit' decisions is remarkably effective as it encourages us to explore options we might otherwise ignore. Recent work suggests that people may actually use such `random exploration' in practice, increasing their behavioral variability when it is more valuable to explore. 
%Other work, however, suggests that variability in human behavior is not always random, and is instead driven by deterministic factors (such as repeating past actions) 
A key question is whether the variability in random exploration is actually random. That is, is random exploration driven by stochastic processes in the brain or by some unobserved deterministic process that we have failed to account for when measuring behavioral variability? By designing an explore-exploit task in which, unbeknownst to them, participants are presented with the exact same choice twice, we provide a partial answer to this question. \added[id=siyu]{By a novel bayesian analysis, we are able to quantify the relative magnitudes of stimulus-driven deterministic noise and non-stimulus-driven random noise in driving random exploration.} In particular, we find evidence that at least 20$\%$ of the variability in random exploration can \added[id=siyu]{indeed} be accounted for by deterministic processing of the stimulus. This still leaves open the possibility that much of random exploration is truly `random', but narrows the window of opportunity for \added[id=siyu]{true} stochastic processes to explain this behavior. \added[id=siyu]{Our work sheds light on how random exploration may be implemented by reducing the filtering of task-irrelevant stimuli in the decision process.}

%suggests that at least some `random' exploration is not random, but leaves open the possibility that the 
%However, it is unclear whether this variability is truly random
%From a modeling perspective, behavioral variability is essentially the variance that can not be explained by a model and is modeled as the level of decision noise. However, what we have called "decision noise" in previous researches could actually just be missing deterministic components from the model, it is difficult to tell whether decision noise truly arises from a stochastic process. Here we show that, while both random and deterministic noise drive variability in behavior, the noise driving random exploration is predominantly random. This suggests that random exploration depends on adaptive noise processes in the brain which are subject to cognitive control.

@@

	%BOB SAYS: IN GENERAL, NEED TO GO THROUGH THE PAPER AND MAKE SURE THAT `RANDOM' NOISE IS REPLACES WITH `NOT A DETERMINISTIC FUNCTION OF THE STIMULUS' OR SOMETHING LIKE THAT
	%SIYU Says: We can define "random noise" as "not a deterministic funciton of the stimulus" at the beginning, and use the term "random noise" in the results section. It feels a little lengthy to use "not a deterministic function of the stimulus" in the results section each time. I changed a few of them back to "random noise" in the results section, in places that I think random noise is better. See if you like the changes.

	%(BOB SAYS: TAKE OR LEAVE THIS CHANGE TO DATE, BUT SETTING UP AS A DATE ALLOWS YOU TO TALK ABOUT AVOID YOUR EX LATER ON, WHICH MAY BE A BETTER EXAMPLE OF A DETERMINISTIC PROCESS!).
	% Siyu Say: I kept this.

	Imagine trying to decide where to go to dinner on a date, you can go to your favorite restaurant, the one you both really enjoy and always go to, or you can try a new restaurant that you know nothing about. Such decisions, in which we must choose between a well-known `exploit' option and a lesser known `explore' option, are known as explore-exploit decisions.  From a theoretical perspective, making optimal explore-exploit choices, i.e. choices that maximize long-term reward, is computationally intractable in most cases \citep{eegittins74, eeBasu18}. In part because of this difficulty, there is considerable interest in how humans and animals solve the explore-exploit dilemma in practice \added[id=siyu]{\citep{Mehlhorn15, SCHULZ20197,WILSON202149}}.
%\citep{eeauer02,eegittins79,eekrebs78,eethompson33, eewatkins89, eebridle90,eemeyer95, eebanks97,eefrank09, eesteyvers09, eelee11, eepl12,eezhang13,eedaw06, eepl11, wilson2014}

	One particularly effective strategy for solving the explore-exploit dilemma is choice randomization \citep{eethompson33, eewatkins89, eebridle90} \added[id=siyu]{, also known as random exploration}. In this strategy, \deleted[id=siyu]{the decision process between exploration and exploitation is corrupted by `decision noise', meaning that }high value `exploit' options are not always chosen and exploratory choices are sometimes made by chance. \added[id=siyu]{In modeling terms, random exploration works by adding 'decision noise' to the value of the options such that sub-optimal exploratory options can sometimes have a higher total score (i.e., value + noise) than the exploit option and get chosen.} \replaced[id=siyu]{Such}{In theory, such} random exploration, is surprisingly effective and, if implemented correctly, can come close to optimal performance \added[id=siyu]{in theory}\citep{eethompson33, eebridle90, eeAgrawal11, eeChapelle11}.

	It has recently been shown that humans appear to use random exploration and can increase such decision noise when it is more beneficial to explore \citep{Gershman2018, wilson2014}. In one of these tasks, known as the Horizon Task, the key manipulation is the horizon condition, i.e. the number of decisions remaining for the participant to make. Increasing the horizon makes exploration more valuable as there is more time to use the information gained by exploration to maximize future rewards. For example, if you are leaving town tomorrow (short horizon), you will probably exploit the restaurant you know and love, but if you are in town for a while (long horizon), you would be more likely to explore the new restaurant. Using such a horizon manipulation it has been shown that people's behavior is more variable in long horizons than short horizons, suggesting that they use adaptive decision noise to solve the explore-exploit dilemma \citep{wilson2014}.

	\replaced[id=siyu]{One limitation, however, is that it is difficult to tell whether what we have called 'decision noise' in previous research is truly a random noise.}{One limitation of this previous work, is that it is difficult to tell whether what is measured as decision noise is truly random.} That is, whether behavioral variability is due to intrinsic stochastic processes in the brain or whether it is due to deterministic processes that we \replaced[id=siyu]{failed to observe}{simply have not observed}. \added[id=siyu]{Decision noise as defined in previous researches are more or less a quantification of what's not predictable by the model. A missing deterministic component from the model could give rise to variability in behavior that might be mistakenly taken as a random noise.} For example, in the restaurant example, my usual preference for one restaurant or another may be overruled if I see an ex romantic partner going into one of them. Avoiding an ex is a deterministic process, but if we fail to take ex's presence into account as scientists modeling the decision, then over a series of such decisions where the ex is present or not, we would mistakenly attribute the ensuing `variability' in choice to randomness.

	%because noise in previous work is defined as the component of the decision process that cannot be accounted for by a model. 
	%, 
	%For example, in the restaurant case above, if you happen to spot an old friend walking in to one of the restaurants, you may be more likely to follow them in. Such a process would be deterministic, in that it is driven by an observable stimulus, but unless a model of your behavior incorporated the presence of your friend,

@@

	In this paper, we investigate the extent to which the apparent randomness in random exploration can be explained by such \added[id=siyu]{unobserved} deterministic processing of the stimulus \added[id=siyu]{(will refer to as deterministic noise)}. \replaced[id=siyu]{To distinguish between stimulus-driven `deterministic noise' and non-stimulus-driven `random noise'}{In particular}, we modify the Horizon Task \citep{wilson2014} to have people face the exact same explore-exploit choice twice. If the decision is a purely deterministic function of the stimulus \added[id=siyu]{(i.e., decision noise is purely deterministic noise)}, then people's choices should be identical both times. That is, their choices should be consistent, since the stimulus is the same both times. Conversely, the more their decision is driven by \replaced[id=siyu]{random noise}{other processes, including both stochastic and unobserved deterministic processes}, the less consistent their behavior should be. By analyzing behavior on this task in both a model-free and model-based manner, we show that at least some of the `randomness' in random exploration must come from deterministic processing of the stimulus. \replaced[id=siyu]{Our work}{This does not prove that random exploration is entirely deterministic or stochastic, but} provides a lower bound on how much deterministic processes contribute to random exploration\added[id=siyu]{, and an upper bound on how much true random noises contribute to random exploration}. \deleted[id=siyu]{This in turn sheds light on how random exploration may be implemented by amplifying the effects of task-irrelevant stimuli on choice.}

	% SIYU says: "Conversely, the more their decision is driven by other processes, including both stochastic and unobserved deterministic processes, the less consistent their behavior should be. " is not correct. Unobserved deterministic processes will be the same for repeated games and will not contribute to choice inconsistency.

@@

@@

	\subsection*{The Repeated-Games Horizon Task}
	We used a modified version of the `Horizon Task' \citep{wilson2014} to show the influence of stimulus-driven `deterministic noise' vs non-stimulus-driven `random noise' \replaced[id=siyu]{in explore-exploit}{on people's} decisions (Figure \ref{fig:taskfig}). In this task, participants make repeated choices between two slot machines, or `one-armed bandits,' that pay out probabilistic rewards. Because they are initially unsure as to the mean payoff of each bandit, this task requires that participants carefully balance exploration of the lesser known bandit with exploitation of the better known bandit to maximize their overall rewards.

	\begin{figure}[hp]
		\begin{center}
			\includegraphics[width=\textwidth]{figures/taskfiga.pdf}
			\caption{ 
				Schematic of the experiment. (A) Dynamics of an example horizon 6 game.  Here the first four trials are forced trials in which participants are instructed which option to play.  After the forced trials, participants are free to choose between the two options for the remainder of the game.  (B) Example repeated games over the course of the experiment.
				% Different possible states of the game after the first free choice over the course of the experiment. 
				On average, participants play more than 150 such games, with varying horizon (1 vs 6), uncertainty condition ([1 3] vs [2 2]) and observed rewards.  In addition, all games are repeated (as Game 18 and 100 are here) such that participants will be faced with the exact same pattern of forced trials and exact same outcomes from those forced trials twice within each experiment.  These repeated games allow us to compute the relative contribution of deterministic and random noise by analyzing the extent to which choices are {\em consistent} across the repeated games.}
			\label{fig:taskfig}
		\end{center}
	\end{figure}

@@

	Relative to the original Horizon Task, the key modification here is to give people `repeated games,' in which they see exact same set of example plays twice in two separate games (separated by several minutes in time so as to avoid detection). By repeating the instructed plays for each game twice, we can set up a situation where (unbeknownst to the participants) they are faced with the exact same explore-exploit choice, with the exact same stimuli twice. Thus, if their behavior is a deterministic function of the stimuli, they will behave identically on the two games, that is their \replaced[id=siyu]{choice}{behavior} on the two versions of each game will be consistent. Conversely, if their behavior is not driven by a deterministic function of the stimulus, then their choices on the repeated games \replaced[id=siyu]{could}{should} be inconsistent. \replaced[id=siyu]{The extent to which participants' choices are consistent on the repeated versions of the games allow}{On average participants played 67.31 repeated games each allowing} us to quantify the extent to which their behavior was \replaced[id=siyu]{driven by a deterministic noise process vs a random noise process}{a deterministic function of the stimulus or not}.

@@

	Before discussing the results for repeated games, we first confirm that the basic behavior in this task is consistent with our previously reported results \citep{wilson2014}. As in our previous work, we find evidence for two types of exploration in the Horizon Task: Random exploration, which is the main focus of this paper, where exploration is driven by noise, and directed exploration, where exploration is driven by information (Figure \ref{fig:modelfree}, Supplementary Figure S1).

	Random exploration is quantified in a model-free way as the probability of choosing the low mean option, $p(\mbox{low mean})$ in the equal, or [2 2], condition. This value increases with horizon, consistent with the idea that behavior is more random in horizon 6 (t(64) = 6.55, p $<$ 0.001 for [1 3], t(64) = 7.99, p $<$ 0.001 for [2 2]).  Directed exploration, is measured as the probability of choosing the more informative option $p(\mbox{high info})$ in the unequal, or [1 3], condition. Again this measure increases with horizon, showing that people are more information seeking in horizon 6 (t(64) = 6.92, p $<$ 0.001).

@@

	Next we asked whether participants' choices were consistent or inconsistent in the two repetitions of each game. The idea behind this measure is that purely deterministic noise should lead to consistent choices as the deterministic stimulus is identical both times. \deleted[id=siyu]{Conversely, if choice is not a deterministic function of the stimulus, participants' choices should be independent, and hence more inconsistent across the repetitions of the game.}\added[id=siyu]{Conversely, if choice is not entirely driven by a deterministic process and is also driven by random noise, participants' choices should be more inconsistent across the repetitions of the game. Moreover, if decision noise is purely random noise, meaning there is no unobserved deterministic process, we will show that we can actually predict the expected level of choice inconsistencies across repetitions of games by accounting for the known deterministic processes and assuming that the random noise process is independent in repetitions of the game.}

	To quantify choice inconsistency we computed the frequency with which participants made different responses for pairs of repeated games (Figure \ref{fig:mf2}, Supplementary Figure S2). Using this measure we found that participants made inconsistent choices in both the unequal ([1 3]) and equal ([2 2]) information conditions, suggesting that not all of the noise was stimulus driven (t-test vs zero revealed that inconsistency was greater than zero for all horizon and uncertainty conditions\deleted[id=siyu]{For [1 3] condition, t(64) = 13.72, p $<$ 0.001 for horizon 1, t(64) = 16.71, p $<$ 0.001 for horizon 6; For [2 2] condition, t(64) = 9.55, p $<$ 0.001 for horizon 1, t(64) = 17.93, p $<$ 0.001 for horizon 6}). In addition, we found that choice inconsistency was higher in horizon 6 than in horizon 1 for both [1 3] and [2 2] condition (For [1 3] condition, t(64) = 5.41, p $<$ 0.001; for [2 2] condition, t(64) = 6.26, p $<$ 0.001), suggesting that at least some of the horizon dependent noise is not a deterministic function of the stimulus\added[id=siyu]{, but rather random noise}.

	\begin{figure}[H]
		\begin{center}
			\includegraphics[width=\textwidth]{figures/RanDetNoise_pinconsistent.jpg}
			\caption{Model-free analysis suggests that both deterministic and random noise contribute to the choice variability in random exploration. For both the [1 3] (A) and [2 2] (B) condition, people show greater choice inconsistency in horizon 6 than horizon 1. However, the extent to which their choices are inconsistent lies between what is predicted by purely deterministic and random noise, suggesting that both noise sources influence the decision.}
			\label{fig:mf2}
		\end{center}
	\end{figure}
